---
title: "Predicting Voter Turnout with tidymodels"
subtitle: "A Modern Machine Learning Ecosystem in R"
author: "Mustafa Salman"
date: today
format:
  beamer:
    theme: metropolis
    colortheme: default
    fonttheme: structurebold
    slide-level: 2
    keep-tex: false
    fig-width: 6
    fig-height: 3.5
    fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidymodels)
library(tidyverse)
library(vip)
library(knitr)
library(palmerpenguins)

set.seed(123)
turnout <- tibble(
  county        = paste("County", 1:500),
  median_income = rnorm(500, 55000, 12000),
  pct_college   = runif(500, 0.15, 0.65),
  pop_density   = rexp(500, rate = 1/500),
  pct_white     = runif(500, 0.3, 0.95),
  region        = sample(c("South","Midwest","Northeast","West"),
                         500, replace = TRUE),
  turnout_pct   = 0.3 + 0.0000045*median_income +
                  0.35*pct_college + rnorm(500, 0, 0.06)
) %>%
  mutate(
    high_turnout = factor(ifelse(turnout_pct > median(turnout_pct),
                                 "High", "Low")),
    region = factor(region)
  )

set.seed(42)
pen_split <- initial_split(drop_na(penguins), prop = 0.75,
                            strata = species)
pen_train <- training(pen_split)
pen_test  <- testing(pen_split)

pen_fit <- workflow() %>%
  add_recipe(
    recipe(species ~ bill_length_mm + bill_depth_mm +
             flipper_length_mm + body_mass_g,
           data = pen_train) %>%
      step_normalize(all_numeric_predictors())
  ) %>%
  add_model(multinom_reg() %>% set_engine("nnet")) %>%
  fit(data = pen_train)

pen_preds <- augment(pen_fit, new_data = pen_test)

set.seed(42)
to_split <- initial_split(turnout, prop = 0.75, strata = high_turnout)
to_train <- training(to_split)
to_test  <- testing(to_split)

to_fit <- workflow() %>%
  add_recipe(
    recipe(high_turnout ~ median_income + pct_college +
             pop_density + pct_white + region,
           data = to_train) %>%
      step_normalize(all_numeric_predictors()) %>%
      step_dummy(all_nominal_predictors())
  ) %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  fit(data = to_train)

to_preds <- augment(to_fit, new_data = to_test)
```

## Who votes — and can we predict it?

- ~60% of eligible Americans voted in 2020, but turnout varies a lot by county
- Campaigns and policymakers want to know: who shows up?
- We have covered regression and inference — tidymodels asks a different question: how accurately can we *classify* counties?

## From inference to prediction

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{What we have done}
\begin{itemize}
  \item Regression: why does turnout vary?
  \item Inference: is income significant?
  \item Bootstrapping: how certain are we?
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{What tidymodels adds}
\begin{itemize}
  \item How accurately can we classify?
  \item Honest train/test evaluation
  \item One workflow for any model
\end{itemize}
\end{column}
\end{columns}

\vspace{0.5em}
\begin{block}{}
\texttt{tidymodels} is to machine learning what \texttt{tidyverse} is to data wrangling.
\end{block}

## What is tidymodels?

A family of packages, one grammar for the whole modeling process:

| Package | Role |
|------------|--------------------------|
| `rsample` | Train/test splits |
| `recipes` | Preprocessing |
| `parsnip` | Model specification |
| `workflows` | Bundle recipe + model |
| `yardstick` | Evaluate performance |

Same tidy principles — just applied to modeling.

## The tidymodels workflow

```{r workflow-diagram, echo=FALSE, fig.height=2, fig.width=7}
nodes  <- tibble(label = c("Split","Recipe","Model","Workflow","Evaluate"),
                 x = 1:5, y = 1)
arrows <- tibble(x = 1:4 + 0.3, xend = 2:5 - 0.3, y = 1, yend = 1)

ggplot() +
  geom_segment(data = arrows,
               aes(x=x, xend=xend, y=y, yend=yend),
               arrow = arrow(length = unit(0.25,"cm"), type="closed"),
               color = "#e94560", linewidth = 0.8) +
  geom_point(data = nodes, aes(x, y), size = 13, color = "#0f3460") +
  geom_text(data = nodes, aes(x, y, label = label),
            color = "white", fontface = "bold", size = 3.2) +
  theme_void() +
  xlim(0.5, 5.5) + ylim(0.5, 1.5)
```

Each step is modular — swap the model without rewriting anything else.

## Vignette: penguin species classification

*Official tidymodels tutorial — predicting species from physical measurements*

```{r pen-code, echo=TRUE, eval=FALSE}
splits <- initial_split(penguins, prop = 0.75,
                        strata = species)
train  <- training(splits)
test   <- testing(splits)

rec <- recipe(species ~ bill_length_mm +
  bill_depth_mm + flipper_length_mm +
  body_mass_g, data = train) %>%
  step_normalize(all_numeric_predictors())

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(multinom_reg() %>%
              set_engine("nnet")) %>%
  fit(data = train)
```

## Vignette results — 97.6% accuracy

```{r pen-plot, fig.height=3.2, fig.width=5.5}
conf_mat(pen_preds, truth = species,
         estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "#d6e4f7", high = "#0f3460") +
  labs(title    = "Confusion Matrix: Penguin Species",
       subtitle = "97.6% accuracy on test set") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## My application: voter turnout

Can county-level socioeconomic data predict whether a county has high or low turnout?

| Predictor | Description |
|----------------|-------------------------------|
| `median_income` | County median household income |
| `pct_college` | % adults with college degree |
| `pop_density` | Population per sq. mile |
| `pct_white` | Racial composition proxy |
| `region` | US Census region |

Outcome: `high_turnout` — above/below county median (binary)

## Building the model

```{r to-code, echo=TRUE, eval=FALSE}
splits <- initial_split(turnout, prop = 0.75,
                        strata = high_turnout)

rec <- recipe(
  high_turnout ~ median_income + pct_college +
    pop_density + pct_white + region,
  data = train
) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(logistic_reg() %>%
              set_engine("glm")) %>%
  fit(data = train)
```

## Results: 77% accuracy, AUC 0.88

```{r roc-plot, fig.height=3.2, fig.width=5.5}
roc_curve(to_preds, truth = high_turnout,
          .pred_High) %>%
  autoplot() +
  labs(title    = "ROC Curve: Voter Turnout",
       subtitle = "AUC = 0.88  |  Accuracy = 77%") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## What predicts turnout?

```{r vip-plot, fig.height=3.2, fig.width=5.5}
to_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 5,
      aesthetics = list(fill = "#533483", alpha = 0.85)) +
  labs(title    = "Variable Importance: Voter Turnout",
       subtitle = "Education and income dominate",
       x = "Importance", y = "Predictor") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## Confusion matrix

```{r conf-plot, fig.height=3.2, fig.width=5.5}
conf_mat(to_preds, truth = high_turnout,
         estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "#e8d5f5", high = "#533483") +
  labs(title    = "Confusion Matrix: Voter Turnout",
       subtitle = "How often did the model get it right?") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## Takeaways

**What we found**

- Education and income are the strongest predictors — consistent with the political science literature

**Why tidymodels?**

- One workflow for any model — swap `logistic_reg()` for `rand_forest()` in one line
- Train/test split gives an honest accuracy estimate
- Natural extension of regression and inference from class

**Limitations**

- Ecological fallacy: county averages are not individual behavior
- Simulated data — real application would use MIT Election Lab

## Thank you

- tidymodels documentation: tidymodels.org
- MIT Election Lab: dataverse.harvard.edu
- GitHub: github.com/msalman-06/tidymodels-voter-turnout

```r

```
