---
title: "Vignette 2: Predicting Voter Turnout with tidymodels"
author: "Mustafa Salman"
date: today
format: html
execute:
  warning: false
  message: false
---

Applying the tidymodels workflow to a new problem: predicting whether a U.S. county has high or low voter turnout based on socioeconomic data. This is a binary classification problem — the outcome has two levels (High/Low) rather than a continuous number. Data is simulated to mirror the structure of the MIT Election Lab county-level dataset (https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ).

```{r}
library(tidymodels) # core modeling ecosystem
library(tidyverse)  # data wrangling and ggplot
library(vip)        # variable importance plots
library(knitr)      # kable() for clean tables
```

```{r}
set.seed(123) # makes the random simulation reproducible

turnout <- tibble(
  county        = paste("County", 1:500),
  median_income = rnorm(500, 55000, 12000),  # normally distributed around $55k
  pct_college   = runif(500, 0.15, 0.65),    # uniform between 15% and 65%
  pop_density   = rexp(500, rate = 1/500),   # right-skewed — most counties are sparse
  pct_white     = runif(500, 0.3, 0.95),
  region        = sample(c("South","Midwest","Northeast","West"),
                         500, replace = TRUE),
  # turnout is modeled as a function of income and education
  # this matches what the political science literature actually finds
  turnout_pct   = 0.3 + 0.0000045*median_income +
                  0.35*pct_college + rnorm(500, 0, 0.06)
) %>%
  mutate(
    # convert continuous turnout to binary: High if above median, Low if below
    # this is the outcome we're trying to predict
    high_turnout = factor(ifelse(turnout_pct > median(turnout_pct),
                                 "High", "Low")),
    region = factor(region) # needs to be a factor for step_dummy() later
  )
```

```{r}
# boxplots by class — checking whether the predictors actually separate High from Low
# if the boxes don't overlap much, the model should work well
turnout %>%
  select(high_turnout, median_income, pct_college) %>%
  pivot_longer(-high_turnout) %>%
  ggplot(aes(x = high_turnout, y = value, fill = high_turnout)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~name, scales = "free_y") +
  scale_fill_manual(values = c("High" = "#533483", "Low" = "#a78bca")) +
  theme_minimal()
```

Counties with high turnout have higher income and more college education, which is consistent with the political science literature.

```{r}
set.seed(42)

# split 75% training, 25% testing
# strata = high_turnout ensures both splits have roughly equal High/Low proportions
# without stratifying, random chance could give us an imbalanced split
splits   <- initial_split(turnout, prop = 0.75, strata = high_turnout)
to_train <- training(splits)
to_test  <- testing(splits)
```

```{r}
# a recipe defines all preprocessing steps before modeling
to_rec <- recipe(
  high_turnout ~ median_income + pct_college +
    pop_density + pct_white + region,
  data = to_train
) %>%
  # normalize so income (in the 50,000s) doesn't dominate predictors on a 0-1 scale
  # sets each numeric predictor to mean = 0, sd = 1
  step_normalize(all_numeric_predictors()) %>%
  # convert region (a factor) into dummy variables
  # models need numbers — this creates binary columns like region_South, region_West
  step_dummy(all_nominal_predictors())
```

```{r}
# logistic regression is the right choice for binary classification
# it predicts the probability of "High" and classifies above/below 0.5
# set_engine("glm") uses base R's glm() function under the hood
to_fit <- workflow() %>%
  add_recipe(to_rec) %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  fit(data = to_train) # fit only on training data — test set stays hidden until eval
```

```{r}
# augment() runs predictions on the test set and appends them to the data
# .pred_class = predicted label, .pred_High = predicted probability of High
to_preds <- augment(to_fit, new_data = to_test)

# accuracy = % of counties correctly classified
# roc_auc = how well the model separates High from Low at all thresholds
# 1.0 is perfect, 0.5 is no better than random guessing
bind_rows(
  accuracy(to_preds, truth = high_turnout, estimate = .pred_class),
  roc_auc(to_preds,  truth = high_turnout, .pred_High)
) %>%
  select(.metric, .estimate) %>%
  kable(digits = 3)
```

```{r}
# ROC curve plots sensitivity vs 1-specificity at every possible threshold
# a good model hugs the top-left corner — the dotted diagonal is random chance
roc_curve(to_preds, truth = high_turnout, .pred_High) %>%
  autoplot() +
  labs(title    = "ROC Curve: Voter Turnout",
       subtitle = "AUC = 0.88") +
  theme_minimal()
```

```{r}
# confusion matrix — rows are predictions, columns are actual values
# diagonal = correct, off-diagonal = errors
# darker = higher count
conf_mat(to_preds, truth = high_turnout,
         estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "#e8d5f5", high = "#533483") +
  theme_minimal()
```

```{r}
# variable importance — which predictors contributed most to the model's decisions
# extract_fit_parsnip() pulls the underlying glm object out of the workflow
to_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 5,
      aesthetics = list(fill = "#533483", alpha = 0.85)) +
  labs(title = "Variable Importance: Voter Turnout") +
  theme_minimal()
```

77% accuracy and AUC of 0.88. Education and income are the strongest predictors which makes sense. Main limitation is ecological fallacy — county averages don't tell us about individual voters.

```{r}
citation("tidymodels")
citation("vip")
```
