---
title: "Vignette 1: Build a Model with tidymodels"
author: "Mustafa Salman"
date: today
format: html
execute:
  warning: false
  message: false
---

Following the official tidymodels "Build a Model" tutorial at https://www.tidymodels.org/start/models/

The data is from Constable (1993) and looks at how three feeding regimes affect the suture width of sea urchins over time.

```{r}
library(tidymodels) # umbrella package — loads parsnip, recipes, rsample, yardstick, etc.
library(readr)      # for read_csv
library(broom)      # tidy() turns model output into a clean tibble
library(dotwhisker) # dwplot() makes coefficient plots
```

```{r}
urchins <- read_csv("https://tidymodels.org/start/models/urchins.csv") |>
  # rename columns to something readable
  setNames(c("food_regime", "initial_volume", "width")) |>
  # food_regime needs to be a factor so R treats it as categorical
  # order matters here — "Initial" is the reference level
  mutate(food_regime = factor(food_regime,
                              levels = c("Initial", "Low", "High")))
```

```{r}
# plot initial_volume vs width, one regression line per feeding regime
# if the slopes look different by group, we need an interaction term
ggplot(urchins, aes(x = initial_volume, y = width,
                    color = food_regime, group = food_regime)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

The slopes look different by feeding regime so an interaction term makes sense.

```{r}
# in tidymodels you specify the model type and engine separately
# this means you can swap engines (e.g. "stan" for Bayesian) without
# changing anything else in the workflow
lm_mod <- linear_reg() |>
  set_engine("lm") # use base R's lm() function to fit

# fit the model — * creates an interaction between initial_volume and food_regime
# meaning the slope of initial_volume is allowed to differ by food_regime
lm_fit <- lm_mod |>
  fit(width ~ initial_volume * food_regime, data = urchins)

# tidy() gives us a clean tibble of coefficients instead of the messy base R output
# estimate = effect size, p.value = same interpretation as in ModernDive
tidy(lm_fit)
```

```{r}
# dwplot makes a coefficient plot — dots are estimates, whiskers are confidence intervals
# the dashed line at 0 is the null hypothesis (no effect)
# anything clearly to the right = positive effect on width
tidy(lm_fit) |>
  dwplot(
    dot_args     = list(size = 2, color = "#0f3460"),
    whisker_args = list(color = "#0f3460"),
    vline = geom_vline(xintercept = 0, linetype = "dashed",
                       color = "gray50")
  ) +
  theme_minimal()
```

```{r}
# generate new data to predict from — one row per food regime, all at volume = 20
new_points <- expand.grid(
  initial_volume = 20,
  food_regime    = c("Initial", "Low", "High")
)

# predict() returns the mean predicted width for each new point
mean_pred <- predict(lm_fit, new_data = new_points)

# type = "conf_int" gives us upper and lower bounds — same idea as
# the confidence intervals we built with bootstrapping in class
conf_int <- predict(lm_fit, new_data = new_points, type = "conf_int")

# bind everything together into one tidy tibble for plotting
plot_data <- new_points |>
  bind_cols(mean_pred) |>
  bind_cols(conf_int)

ggplot(plot_data, aes(x = food_regime)) +
  geom_point(aes(y = .pred), size = 3, color = "#0f3460") +
  geom_errorbar(aes(ymin = .pred_lower, ymax = .pred_upper),
                width = 0.15, color = "#0f3460") +
  labs(x = "Food Regime", y = "Predicted Width") +
  theme_minimal()
```

Urchins in the High feeding regime are predicted to be wider at 20mL. The confidence intervals for Initial and Low overlap, so we can't distinguish those two clearly.

The main thing tidymodels adds here over base R: swapping `set_engine("lm")` for `set_engine("stan")` fits a Bayesian model with the exact same code — `fit()`, `predict()`, and `tidy()` all work identically regardless of engine.

```{r}
citation("tidymodels")
```
